Sequential rank agreement methods for comparison of ranked lists

### 2015-06-15 latex/paper.Rnw
Tekst

\title{Sequential rank agreement methods for comparison of ranked
  lists}
\author{Claus Thorn Ekstrøm}
\author{Thomas Alexander Gerds}
\author{Kasper Brink-Jensen}
\affil{Biostatistics, University of Copenhagen}

@@

<<echo=FALSE>>=
#note: always pass alpha on the 0-255 scale
makeTransparent<-function(someColor, alpha=100)
{
  newColor<-col2rgb(someColor)
  apply(newColor, 2, function(curcoldata){rgb(red=curcoldata[1], green=curcoldata[2],
    blue=curcoldata[3],alpha=alpha, maxColorValue=255)})
}
@ %

@@

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found in the $L$ top $d$ parts of the lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
and small values close to zero suggests that the lists agree on the
ordering while larger values suggests disagreement. If the ranked
lists are identical then the value of SRA will be zero for all depths
$d$.  The sequential rank agreement can be interpreted as the average
distance of the individual rankings of the lists from the average
ranking for each of the items we have seen until depth $d$.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$.  An example is seen in the
left panel of Figure~\ref{fig:example1} where four analysis methods
were used to rank genes (for further detail see the application
section below). The sequential rank agreement at depth $d$ can be
interpreted as how many ranks apart do we on average expect the items
found until depth $d$ to be so smaller values on the $y$ axis
correspond to better agreement among the lists than larger values.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig1a}
\includegraphics[width=.45\textwidth]{paper-fig1b}
\end{center}
 \caption{Left panel: Sequential rank agreement for 4 different analysis methods
   applied to the 3051 genes in the Golub data (solid line). The dashed line shows the
   corresponding estimated sequential rank agreement for censored
   lists based purely on the top 50 ranked genes. Right panel contains
   the same information but has zoomed in on the first part of the $x$-axis.}
 \label{fig:example1}
\end{figure}

@@

producelists <- function(x, y) {
    nitems <- nrow(x)
    index <- seq(1, nrow(golub))
                                        # d <- data.frame(y, x)

                                        # Marginale t-tests
    mt.p <- sapply(index, function(i) { t.test(x[i,] ~ y)$p.value } )
    list1 <- order(mt.p)

                                        # Marginale logreg-tests
    mlogreg.p <- sapply(index, function(i) { drop1(glm(y ~ x[i,], family=binomial), test="Chisq")[2,5] } )
    list2 <- order(mlogreg.p)

                                        # Elastic net
    X <- scale(t(x))
    enet <- glmnet(X, y, family="binomial", alpha=.8)
    nyres <- cv.glmnet(X, y, family="binomial", alpha=.8)
    coefficients <- coef(enet, s=nyres$lambda.1se)[-1]
    nonzeros <- sum(coefficients!=0)
    list3 <- order(abs(coefficients), decreasing=TRUE)
    if (nonzeros<nitems) {
        list3[(nonzeros+1):nitems] <- NA
    }
                                        # MIC
    MIC <- sapply(index, function(i) { mine(x[i,], y)$MIC})
    list4 <- order(MIC, decreasing=TRUE)

inputmatrix <- producelists(x, y)
colnames(inputmatrix) <- c("T", "LogReg", "ElasticNet", "MIC", "RF")[1:ncol(inputmatrix)]

res <- sqrt(sra(inputmatrix, B=20))
inputmatrix2 <- inputmatrix
inputmatrix2[51:nrow(inputmatrix),] <- NA
res2 <- sqrt(sra(inputmatrix2, B=20))

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE>>=
ysize <- 1000
plot(res[1:ysize], lwd=3, col="red", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")
lines(res2[1:ysize], lwd=2, col="purple")

<<fig1b, fig=TRUE,echo=FALSE,include=FALSE >>=
ysize <- 80
plot(res[1:ysize], lwd=3, col="red", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")
lines(res2[1:ysize], lwd=2, col="purple")

<<fig2a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE>>=
ysize <- 1000
plot(res[1:ysize], lwd=3, col="red", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")
lines(res2[1:ysize], lwd=2, col="purple")
##
# col1 <- makeTransparent()
null <- sqrt(random_list_sra(inputmatrix, B=1, n=40))
null3 <- sqrt(random_list_sra(inputmatrix, B=20, n=20))
## matlines(null[1:ysize,], col="gray", lty=1)

##

##null2 <- sapply(1:40, function(i) {
##    sra(producelists(x, sample(y)), B=1)
##} )
## matlines(null2[1:ysize,], col="lightblue", lty=1)

load("R/fig1null.rda")

bcolor <- makeTransparent("lightblue", alpha=80)
bcolor2 <- makeTransparent("red", alpha=80)
bcolor3 <- makeTransparent("green", alpha=80)

www <- smooth_sra(null2)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor, border=NA)

www <- smooth_sra(null)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor2, border=NA)

www <- smooth_sra(null3)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor3, border=NA)

<<fig2b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE >>=
ysize <- 80
plot(res[1:ysize], lwd=3, col="red", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")
lines(res2[1:ysize], lwd=2, col="purple")
##
# col1 <- makeTransparent()
null <- sqrt(random_list_sra(inputmatrix, B=1, n=40))
null3 <- sqrt(random_list_sra(inputmatrix, B=20, n=20))
## matlines(null[1:ysize,], col="gray", lty=1)

##

##null2 <- sapply(1:40, function(i) {
##    sra(producelists(x, sample(y)), B=1)
##} )
## matlines(null2[1:ysize,], col="lightblue", lty=1)

load("R/fig1null.rda")

bcolor <- makeTransparent("lightblue", alpha=80)
bcolor2 <- makeTransparent("red", alpha=80)
bcolor3 <- makeTransparent("green", alpha=80)

www <- smooth_sra(null2)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor, border=NA)

www <- smooth_sra(null)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor2, border=NA)

www <- smooth_sra(null3)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor3, border=NA)

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig2a}
\includegraphics[width=.45\textwidth]{paper-fig2b}
\end{center}
 \caption{Sequential rank agreement for 4 different analysis methods
   applied to the Golub data (thick red line). The red and blue shaded
 areas represent the pointwise 95\% normal areas for the XXX and XXY benchmarks,
respectively.}
 \label{fig:case1}
\end{figure}

<<label=tab1,echo=FALSE,results=tex>>=
library(xtable)
xxx <- as.data.frame(inputmatrix[1:10,])
xxx2 <- xtable(xxx)
digits(xxx2) <- 0
print(xxx2)
#print(xtable(xxx, caption = "Top 10 list of results", label = "tab:one",
#digits = c(0, 0, 2, 0, 2, 3, 3)), table.placement = "tbp",
#caption.placement = "top")

>>>>

\title{Sequential rank agreement methods for comparison of ranked
  lists}
\author{Claus Thorn Ekstrøm}
\author{Thomas Alexander Gerds}
\author{Andreas Jensen}
\author{Kasper Brink-Jensen}
\affil{Biostatistics, University of Copenhagen}

\renewcommand{\topfraction}{.85}
\renewcommand{\bottomfraction}{.7}
\renewcommand{\textfraction}{.15}
\renewcommand{\floatpagefraction}{.66}
\renewcommand{\dbltopfraction}{.66}
\renewcommand{\dblfloatpagefraction}{.66}
\setcounter{topnumber}{9}
\setcounter{bottomnumber}{9}
\setcounter{totalnumber}{20}
\setcounter{dbltopnumber}{9}

@@

<<echo=FALSE>>=
#note: always pass alpha on the 0-255 scale
makeTransparent<-function(someColor, alpha=100)
{
  newColor<-col2rgb(someColor)
  apply(newColor, 2, function(curcoldata){rgb(red=curcoldata[1], green=curcoldata[2],
    blue=curcoldata[3],alpha=alpha, maxColorValue=255)})
}

@@

We now describe what is exemplified in Panel (c) of Table
\ref{tab:example} and how it can be used to define \emph{sequential
  rank agreement}. For an integer $1\le d\le P$ we define the unique
set of items found in the $L$ top $d$ parts of the lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{\pi_l(r) ; r \leq d, l = 1, \ldots, L \}.
\end{equation}
The \emph{sequential rank agreement} is the pooled standard deviation
of the items found in the set $S_d$:
\begin{equation}
\textrm{SRA}(d)= \sqrt{\frac{\sum_{\{p \in
      S_d\}}(L-1)A(X_p)^2}{(L-1)|S_d|}}, \label{def:sra}
\end{equation}
and small values close to zero suggests that the lists agree on the
ordering while larger values suggests disagreement. If the ranked
lists are identical then the value of SRA will be zero for all depths
$d$.  The sequential rank agreement can be interpreted as the average
distance of the individual rankings of the lists from the average
ranking for each of the items we have seen until depth $d$. Note that
the terms that are part of \eqref{def:sra} are not independent since
each list contains the ranks from 1 tp $P$ exactly once. Hence we
simply use the pooled standard deviation as a measure of the average
distance between rankings.

@@

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to improve the classification of acute leukemias
between two types: acute lymphoblastic leukemia (ALL) or acute myeloid
leukemia (AML) \citep{Golub1999}. Preprocessing of the gene expression
data was done as described in \citet{Dudoit2002} and the four
different analysis approaces were: marginal two-sample $t$ tests,
marginal logistic regression analyses, logistic regression eleastic
net, and marginal maximum information content correlations (MIC)
\citep{Reshef2011} which resulted in the rankings seen in
Table~\ref{tab1}.  The sequential rank agreement at depth $d$ can be
interpreted as how many ranks apart do we on average expect the items
found until depth $d$ to be so smaller values on the $y$ axis
correspond to \emph{better} agreement among the lists than larger
values.

topk <- 20

producelists <- function(x, y) {
    nitems <- nrow(x)
    index <- seq(1, nrow(golub))
    ## d <- data.frame(y, x)

    ## Marginale t-tests
    mt.p <- sapply(index, function(i) { t.test(x[i,] ~ y)$p.value } )
    list1 <- order(mt.p)

    ## Marginale logreg-tests
    mlogreg.p <- sapply(index, function(i) { drop1(glm(y ~ x[i,], family=binomial), test="Chisq")[2,5] } )
    list2 <- order(mlogreg.p)

    ## Elastic net
    X <- scale(t(x))
    enet <- glmnet(X, y, family="binomial", alpha=.8)
    nyres <- cv.glmnet(X, y, family="binomial", alpha=.8)
    coefficients <- coef(enet, s=nyres$lambda.1se)[-1]
    nonzeros <- sum(coefficients!=0)
    list3 <- order(abs(coefficients), decreasing=TRUE)
    if (nonzeros<nitems) {
        list3[(nonzeros+1):nitems] <- NA
    }
    ## MIC
    MIC <- sapply(index, function(i) { mine(x[i,], y)$MIC})
    list4 <- order(MIC, decreasing=TRUE)

## Original run
B <- 20
inputmatrix <- producelists(x, y)
colnames(inputmatrix) <- c("T", "LogReg", "ElasticNet", "MIC", "RF")[1:ncol(inputmatrix)]
res <- sra(inputmatrix, B=B)

## Now make the same run but where we only have top 50 lists
inputmatrix2 <- inputmatrix
inputmatrix2[(topk+1):nrow(inputmatrix),] <- NA
res2 <- sra(inputmatrix2, B=B)

<<fig1a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

## Compute random list backgrounds
null <- random_list_sra(inputmatrix, B=B, n=nreference)

bcolor <- makeTransparent("red", alpha=80)
bcolor2 <- makeTransparent("blue", alpha=80)

## Plot the random lists
www <- smooth_sra(null)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor2, border=NA)

## Load the data for the methods
makereference <- function(x, y, n=100, B=20) {
    sapply(1:n, function(i) { sra(producelists(x, sample(y)), B=B) }  )
}
## Uncomment the two lines below to redu the analysis
###null2 <- makereference(x, y, n=100, B=B)
###save(null2, file="R/fig1null2.rda")
load("R/fig1null2.rda")

## Plot the reference sras
www2 <- smooth_sra(null2)
polygon(c(1:length(www2$lower), rev(1:length(www2$lower))),
        c(www2$lower, rev(www2$upper)),
        col=bcolor, border=NA)

<<fig1b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE >>=
## Same as above but zoomed in
ysize <- 50
plot(res[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

## Plot the random lists
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor2, border=NA)

## Plot the reference sras
www2 <- smooth_sra(null2)
polygon(c(1:length(www2$lower), rev(1:length(www2$lower))),
        c(www2$lower, rev(www2$upper)),
        col=bcolor, border=NA)

<<fig2a, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE>>=
ysize <- 1000
nreference <- 20
## Plot the line
plot(res2[1:ysize], col="black", lwd=3, type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

## Compute random list backgrounds
null <- random_list_sra(inputmatrix2, B=B, n=nreference)

## Plot the random lists
www <- smooth_sra(null)
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor2, border=NA)

## Load the data for the methods
makereference <- function(x, y, n=100, B=20) {
    sapply(1:n, function(i) { mymat <- producelists(x, sample(y)) ; mymat[(topk+1):nrow(mymat),] <- NA ; sra(mymat, B=B) }  )
}
## Uncomment the two lines below to redu the analysis
##null2 <- makereference(x, y, n=100, B=B)
##save(null2, file="R/fig1null2c.rda")
load("R/fig1null2c.rda")

## Plot the reference sras
www2 <- smooth_sra(null2)
polygon(c(1:length(www2$lower), rev(1:length(www2$lower))),
        c(www2$lower, rev(www2$upper)),
        col=bcolor, border=NA)

<<fig2b, fig=TRUE,echo=FALSE,include=FALSE,cache=TRUE >>=
## Same as above but zoomed in
ysize <- 50
plot(res2[1:ysize], lwd=3, col="black", type="l", ylim=c(0,1300), ylab="Sequential rank agreement", xlab="Depth")

## Plot the random lists
polygon(c(1:length(www$lower), rev(1:length(www$lower))),
        c(www$lower, rev(www$upper)),
        col=bcolor2, border=NA)

## Plot the reference sras
www2 <- smooth_sra(null2)
polygon(c(1:length(www2$lower), rev(1:length(www2$lower))),
        c(www2$lower, rev(www2$upper)),
        col=bcolor, border=NA)

<<label=tab1,echo=FALSE,results=tex>>=
library(xtable)
xxx <- as.data.frame(cbind(1:10, inputmatrix[1:10,]))
colnames(xxx) <- c("Ranking", colnames(xxx)[2:5])
xxx2 <- xtable(xxx, caption = "Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method")
digits(xxx2) <- 0
print(xxx2, table.placement = "tb", caption.placement = "top", include.rownames=FALSE)
#print(xtable(xxx,  label = "tab:one",

\begin{figure}[tb]
\begin{center}
\includegraphics[width=.45\textwidth]{paper-fig1a}
\includegraphics[width=.45\textwidth]{paper-fig2a}
\includegraphics[width=.45\textwidth]{paper-fig1b}
\includegraphics[width=.45\textwidth]{paper-fig2b}

\end{center}
 \caption{Left panels: Sequential rank agreement for 4 different analysis methods
   applied to the 3051 genes in the Golub data. The two plots on the
   right show the sequential rank agreement for the same data but
   where only the top-\Sexpr{topk} ranked items are available. The blue
   and red areas correspond to the independent and randomized
   reference hypothesis areas, respectively. The bottom plots are
   identical to the corresponding top plots but have been zoomed in on
   the first part of the $x$ axis.}
 \label{fig:example1}
\end{figure}

@@
